No one-size-fits-all solution to clean GBIF

### 2020-02-29 manuscript_peerj.Rmd
peerj format

---
title: "No one-size-fits-all solution to clean GBIF"
author: 
- Alexander Zizka^1,2^,
- Fernanda Antunes Carvalho^4^,
- Alice Calvente^4^,
- Ana Paula Alves da Silva^4^,
- Andressa Cabral^5^,
- Arthur de Souza Soares^4^,
- Eduardo Calisto Tomaz^4^,
- Fernanda Gondim Lambert Moreira^4^,
- Filipe Cabreirinha Serrano^6^,
- Jéssica Fernanda Ramos Coelho^4^,
- Mabel Rocio Baez-Lizarazo,^7^
- Mariana Ramos Fantinati^4^,
- Matheus Colli-Silva^5^,
- Moabe Ferreira Fernandes^8^,
- Nathália Michelly da Cunha Santos^4^,
- Paolla Gabryelle Cavalcante de Souza^4^,
- Renata Clicia dos Santos‐Costa^4^,
- Thais Ferreira-Araújo^4^,
- Tiago Andrade Borges Santos^9^,
- Tiago Luiz Vieira^9^,
- Valéria Fonseca Vale^4^,
- Alexandre Antonelli^10,11,12^
bibliography: course_natal.bib
csl: jbi-3.csl
output:
  #word_document:
  bookdown::pdf_document2:
    toc: false
    toc_depth: 1
    fig_caption: true
    number_sections: false
    keep_tex: yes
nocite: |
    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2019q
header-includes: |
    \usepackage{caption}
    \usepackage{fancyhdr}
    \usepackage{booktabs}
    \pagestyle{fancy}
    \fancyhead[L]{Automated filter effect}
    \fancyhead[LE,RO]{\thepage}
    \fancyfoot{}
    \usepackage{float}
    \floatplacement{figure}{H}
    \usepackage{lineno}
    \linenumbers
linestretch: 1.3
fontsize: 12pt
always_allow_html: yes
---

1. sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
2. Naturalis Biodiversity Center, Leiden, The Netherlands
3. Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
4. Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
5. Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
6. Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
7. Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
8. Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
9. Universidade Estadual de Feira de Santana, Bahia, Brazil
10. Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
11. Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
12. Royal Botanic Gardens Kew, Richmond, United Kingdom

\newpage{}

@@

# Abstract
Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF. We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect. While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.

# Keywords
Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.

\newpage{}

# Introduction
Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

@@

# Material and Methods
## Choice of study taxa
This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group. Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree in documented expertise and number of years working on each group. We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

## Species occurrence data
<!-- @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o -->

@@

##  Automated cleaning
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e. test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

## Downstream analyses
We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2016]. We then performed an automated conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017]. `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].

## Evaluation of results
<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear. Did people provide a specific score for those, or how were their assessments recorded? I think we need to clarify this, as it’s the weakest part of the paper at the moment. We don’t report the results for this evaluation nor what they mean.  -->
Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.

@@

# Results
```{r}
# Table 1
tab1 <- read_excel("figures_tables/table_study_taxa.xlsx")

tab1 <- tab1 %>%
  mutate_all(linebreak) %>%
  select(-researcher) %>% 
  filter(!taxon %in% c("Entomobryidae", "Opisthonema", "Eugenia", "Harpalyce"))

# Table 2
tab2 <- read_csv("output/summary_table_flags.csv") %>% 
  dplyr::select(-coordinate_missing, -invalid_coords, -sea_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
  mutate(fraction = round(fraction * 100,1))

@@

# Discussion
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

<!-- 1. Which is the most relevant flagging mechanism? -->
Several reasons may explain the high proportion of records flagged as duplicates. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics. Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.

@@

<!-- 2. Does the importance of individual errors differ among taxonomic groups? -->
The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals. Therefore, the same tests can be used across taxonomic groups. Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability. The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.

@@

The accuracy of the automated conservation assessment was in the same range as found by previous studies [@NicLughadha2019; @Zizka2019a]. The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies [@Gueta2016]. The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering. This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF. Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available. The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records [@Rivers2011].

# Conclusions
Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses. While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.

@@

# Acknowledgements
This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização). AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv. AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.

# Supplementary material
Appendix 1 - Analyses scripts

Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test

Appendix 3 - The full results of the conservation assessment

Appendix 4 - Taxon specific richness maps and comments.

\newpage{}

# Tables

<!-- A table presenting the study taxa -->
```{r tabletaxa, warning=FALSE}
kable(tab1, booktabs = T, escape = T, format="latex",
      col.names = linebreak(c("Taxon",
                              "Taxon rank",
                              "Realm",
                              "Common name",
                              "'Phylum'",
                              "Order",
                              "Family")),
        caption = "The study groups and their taxonomy. This study includes three marine and 13 terrestrial taxa, six of them animals, one group of fungi and nine plants, belonging to 16 different orders.")%>%
  landscape()%>%
  kable_styling(font_size = 11) %>%
   row_spec(6, hline_after = TRUE) %>%
   row_spec(7, hline_after = TRUE) %>%
   column_spec(1, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(3, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(4:5, width = "2.5cm", bold = FALSE, italic = FALSE)
```

@@

\newpage{}

# Figures
<!-- A map with number of species and where they have been removed -->
```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 16 Neotropical taxa including plants and animals, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
```

@@

\newpage{}

# References

>>>>

---
title: No one-size-fits-all solution to clean GBIF
preprint: false
author: 
  - name: Alexander Zizka
    affiliation: 1,2
    corresponding: true
    email: alexander.zizka@idiv.de
  - name: Fernanda Antunes Carvalho
    affiliation: 3
  - name: Alice Calvente
    affiliation: 4
  - name: Ana Paula Alves da Silva
    affiliation: 4
  - name: Andressa Cabral
    affiliation: 5
  - name: Arthur de Souza Soares
    affiliation: 4
  - name: Eduardo Calisto Tomaz
    affiliation: 4
  - name: Fernanda Gondim Lambert Moreira
    affiliation: 4
  - name: Filipe Cabreirinha Serrano
    affiliation: 6
  - name: Jéssica Fernanda Ramos Coelho
    affiliation: 4
  - name: Mabel Rocio Baez-Lizarazo
    affiliation: 7
  - name: Mariana Ramos Fantinati
    affiliation: 4
  - name: Matheus Colli-Silva
    affiliation: 5
  - name: Moabe Ferreira Fernandes
    affiliation: 8
  - name: Nathália Michelly da Cunha Santos
    affiliation: 4
  - name: Paolla Gabryelle Cavalcante de Souza
    affiliation: 4
  - name: Renata Clicia dos Santos‐Costa
    affiliation: 4
  - name: Thais Ferreira-Araújo
    affiliation: 4
  - name: Tiago Andrade Borges Santos
    affiliation: 9
  - name: Tiago Luiz Vieira
    affiliation: 9
  - name: Valéria Fonseca Vale
    affiliation: 4
  - name: Alexandre Antonelli
    affiliation: 10,11,12
affiliation:
  - code: 1
    address: sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
  - code: 2
    address: Naturalis Biodiversity Center, Leiden, The Netherlands
  - code: 3
    address: Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
  - code: 4
    address: Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
  - code: 5
    address: Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
  - code: 6
    address: Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
  - code: 7
    address: Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
  - code: 8
    address: Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
  - code: 9
    address: Universidade Estadual de Feira de Santana, Bahia, Brazil
  - code: 10
    address: Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
  - code: 11
    address: Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
  - code: 12
    address: Royal Botanic Gardens Kew, Richmond, United Kingdom
abstract: >
  Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF. We find that 29-67% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect. While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research. 
bibliography: course_natal.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
  rticles::peerj_article: default
nocite: |
    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2019q
---

@@

# Introduction {-}
Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

@@

# Material and Methods {-}
## Choice of study taxa {-}
This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group. Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree in documented expertise and number of years working on each group. We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

## Species occurrence data {-}
<!-- @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o -->

@@

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e. test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

## Downstream analyses {-}
We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2016]. We then performed an automated conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017]. `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].

## Evaluation of results {-}
<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear. Did people provide a specific score for those, or how were their assessments recorded? I think we need to clarify this, as it’s the weakest part of the paper at the moment. We don’t report the results for this evaluation nor what they mean.  -->
Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.

@@

# Results {-}
```{r}
# Table 1
tab1 <- read_excel("figures_tables/table_study_taxa.xlsx")

tab1 <- tab1 %>%
  mutate_all(linebreak) %>%
  select(-researcher) %>% 
  filter(!taxon %in% c("Opisthonema", "Eugenia"))

# Table 2
tab2 <- read_csv("output/summary_table_flags.csv") %>% 
  dplyr::select(-coordinate_missing, -invalid_coords, -gbif_coords,-inst_coords, -equal_coords) %>% 
  mutate(fraction = round(fraction * 100,1))

@@

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

## The relevance of individual filters {-}
Several reasons may explain the high proportion of records flagged as duplicates. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics. Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Duplicated records do not represent errors per se, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.

@@

## Similarities and differences among taxa {-}
The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals. Therefore, the same tests can be used across taxonomic groups. Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability. The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods (See Appendix 4 for the individual groups), which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.

@@

## The impact of filtering on the accuracy of automated conservation assessments {-}
The accuracy of the automated conservation assessment was in the same range as found by previous studies [@NicLughadha2019; @Zizka2019a]. The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies [@Gueta2016]. The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering. This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF. Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available. The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records [@Rivers2011].

# Conclusions {-}
Our results suggest that between one to two thirds of the occurrence records obtained from public databases might be unsuitable for downstream biodiversity analyses. While the majority of these records might not be erroneous per se, they might be overly imprecise and thereby increase uncertainty of downstream results or add computational burden on big data analyses.

@@

# Acknowledgments {-}
This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alexandre Antonelli and Alexander Zizka through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização). AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv. AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew.

# Supplementary material {-}

- Appendix 1 - Analyses scripts

- Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test

- Appendix 3 - The full results of the conservation assessment

- Appendix 4 - Taxon specific richness maps and comments.

# Tables {-}
<!-- A table presenting the study taxa -->
```{r tabletaxa, warning=FALSE}
kable(tab1, booktabs = T, escape = T, format="latex",
      col.names = linebreak(c("Taxon",
                              "Taxon rank",
                              "Realm",
                              "Common name",
                              "'Phylum'",
                              "Order",
                              "Family")),
        caption = "The study groups and their taxonomy. This study includes three marine and 13 terrestrial taxa, six of them animals, one group of fungi and nine plants, belonging to 16 different orders.")%>%
  landscape()%>%
  kable_styling(font_size = 11) %>%
   row_spec(6, hline_after = TRUE) %>%
   row_spec(7, hline_after = TRUE) %>%
   column_spec(1, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(3, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(4:5, width = "2.5cm", bold = FALSE, italic = FALSE)
```

@@

# Figures {-}
<!-- The study species-->
```{r species, out.width="0.9\\linewidth", fig.cap=c('Example of study taxa. \\textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \\textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \\textbf{C)} Tillandsia recurva, EC Tomaz; \\textbf{D} \\textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano;\\textbf{E)} Neanuridae (\\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \\textbf{F)} \\textit{Tocoyena formosa}, Fernanda GL Moreira; \\textbf{G)} \\textit{Harengula jaguana}, Thais Ferreira-Araujo; \\textbf{H)} \\textit{Gaylussacia decipiens}, Luiz Menini Neto; \\textbf{I)} \textit{Ooocephalus foliosus}, Arthur Soares; \\textbf{J)} \\textit{Tityus carvalhoi}, Renata C Santos-Costa; \\textbf{K)} \\textit{Prostheca vespa}, Tiago veira'), echo=FALSE}
knitr::include_graphics(path = "./figures_tables/Figure_study_groups.jpg")
```

<!-- A map with number of species and where they have been removed -->
```{r total, out.width="\\linewidth", fig.cap=c('The difference in records available between the raw and filtered dataset of 16 Neotropical taxa including plants and animals, plotted in a 100x100 km grid across the Neotropics.'), echo=FALSE}
knitr::include_graphics(path = "./output/figure_number_of_records.jpg")
```

@@

\newpage{}
# References {-}
